---
title: "Class 18: Pertussis mini-project"
author: "Anna Waters (PID: A16271985)"
format: pdf
editor: visual
---

First we will examine and explore Pertussis case numbers in the US as tracked by the CDC:https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

We can use the datapasta package to scrape this data from the website into R:

```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```


>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time. 

I want a plot of cases per year:

```{r}
library(ggplot2)

ggplot(cdc, aes(year, cases)) +
  geom_line()

```

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc, aes(year, cases)) +
  geom_line() +
  geom_vline(xintercept = 1946, col= "blue") +
  geom_vline(xintercept = 1996, col = "red") +
  geom_vline(xintercept = 2020, col= "purple")
```

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the aP vaccine for about 10 years, case numbers were low and stable, then there was a large increase in cases until 2020. 2020 and onwards, there was a dramatic decrease in the number of cases. 

Access data from the CMI-PB project:

This database (like many modern projects) uses an API to return JSON format data. 

We will use the R package `jsonlite`.
```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/subject", simplifyVector = T)

head(subject)
```
>How many wP (the older whole-cell vaccine) individuals and aP (the newer acellular vaccine) individuals are in this data base?

```{r}
table(subject$infancy_vac)
```
There are 58 patients with wP and 60 patients with aP vaccines. 

>How many male and female patients?

```{r}
table(subject$biological_sex)
```
79 female patients and 39 male patients

>What is the number of individuals by biological sex and race?

```{r}
table(subject$biological_sex, subject$race)
```
There are a lot more women than men enrolled in the study (2x more). There is also much higher numbers of white and Asian people which does not reflect the demographics of the US over all. 

```{r}
subject$year_of_birth
```
# Side-Note: Working with dates 

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
today() - ymd("2000-01-01")
```

```{r}
time_length(today() - ymd("2002-10-11"), "years")
```

So what is the age of everyone on our dataset:

```{r}
subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")
```

>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(dplyr)
ap <- subject %>% 
  filter(infancy_vac == "aP")

mean(ap$age)
```

```{r}
wp <- subject %>% 
  filter(infancy_vac == "wP")

mean(wp$age)
```



```{r}
ggplot(subject, aes(age, fill = infancy_vac)) +
  geom_histogram() +
  facet_wrap(vars(infancy_vac), nrow = 2)
```

## Get more data from CMI-PB

```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = T)
head(specimen)
```

Merge the two tables together with functions from `dplyr`. 

```{r}
library(dplyr)
```

We need to **join** thse two tables (subject and specimen) to make a single new "meta" table with all our metadata. We will use the `dplyr` package join functions to do this. 

```{r}
meta <- inner_join(subject, specimen)
head(meta)
```

Now we can read some of the other data from CMI-PB

```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = T)

head(ab_titer)
```

One more `inner_join()` to add all our metadata in `meta` on to our `ab_titer` table:

```{r}
abdata <- inner_join(ab_titer, meta)

head(abdata)
```

Our first exploratory plot: 

```{r}
table(abdata$antigen)
```


```{r}
ggplot(abdata, aes(MFI, antigen)) +
  geom_boxplot()
```

>Why are certain antigens and not others very variable in their detected levels here?

Some of these antigens are related to pertussis and others aren't. This data includes antigen counts from before and after vaccination as well.

>Can you facet or even just color by infancy_vac?

```{r}
ggplot(abdata, aes(MFI, antigen, col= infancy_vac)) +
  geom_boxplot() 
  #facet_wrap(vars(infancy_vac))
```

There are potentially aome differences here but in general it is hard to tell with this whole dataset overview...

```{r}
table(abdata$dataset)
```

Lets focus in on just the 2021 dataset:

```{r}
abdata.21 <- filter(abdata, dataset == "2021_dataset")

table(abdata.21$dataset)
```


Focus on PT antigen IcG levels
```{r}
pt.21 <- filter(abdata.21, isotype == "IgG", antigen == "PT")
```


plot of days (time) related to boost vs MFI levels


```{r}
ggplot(pt.21) +
  aes(x=planned_day_relative_to_boost, y=MFI_normalised, col=infancy_vac) +
  geom_point() 
```

```{r}
ggplot(pt.21) +
  aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
      group = subject_id) +
  geom_point() +
  geom_line() +
  labs(title = "2021 dataset IgG PT") +
  geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed")
```

> Q17. Do you see any clear difference in aP vs. wP responses?

There is clearly higher response rates in the wP after the booster compared to the aP patients. This is very exentuated at day 14, where the majority of wP points are high above aP for the PT antigent. 