---
title: "class16:graphing the tsv"
author: "Anna Waters (PID: A16271985)"
format: pdf
editor: visual
---

Reading in file from virtual computer

```{r}
data <- read.delim("mm-second.x.zebrafish.tsv", sep = "\t")
```
 
 Renaming columns
 
```{r}
colnames(data) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
head(data)
```
 
```{r}
hist(data$bitscore, breaks = 30)
```
The data is not very normal, it is somewhat skewed towards lower bitscores. There are peaks around 250 and 50 making it somewhat binomial.  

Making the tsv file plots
```{r}
library(ggplot2)
ggplot(data, aes(pident, bitscore)) + geom_point(alpha=0.1) +
  xlab("Percent Identity") +
  ylab("Bitscore")
```

Looks like percent identiy and bitscore are not completely correlated with each other. 

```{r}
ggplot(data, aes((data$pident * (data$qend - data$qstart)), bitscore)) + geom_point(alpha=0.1) + geom_smooth() +
  xlab("Percent Identity x Length") +
  ylab("Bitscore")
```
This graph shows that the correlation between the percent identity and sequence length with the bitscores is very high and positive. 
